### 2010-03-30 papers/icfp-2010-desc/paper_idesc.tex
Some suggestions to fix some typos

However, this new class of objects cannot be defined by mere
induction. In the case of vectors, for instance, we have to define the
whole \emph{family} of vector in one go: vectors of all size needs to
be defined at the same time. Where simply-typed languages have
inductive types, in dependently typed programming we the basic grammar of
data-types should be that of inductive families of types. To this end, we
rely on \emph{indexing}. We are going to transform the $\Desc$ universe into
$\IDesc{\!}$, the universe of indexed descriptions. We express inductive families
by the $\V{I} \To \IDesc{\V{I}}$ type.  \note{pwm: Vec might not be the best
motivating example here, since it doesn't appear in the paper (more on that
later). Come to think of it, this is entirely too vague without spelling out
a concrete example.}

\subsection{Desc, atomically}

@@

Before moving on indexed descriptions, we have to carry out some
maintenance work on descriptions. We presented $\Desc$ as the grammar
of inductive types. Hence, the codes closely follow this grammar. In
the following, we adopt an alternative presentation. With
$\DSigma{\!}{\!}$, we are actually \emph{quoting} a standard
type-former, namely

@@

\[\stk{
\data \IDesc{\!} (\Bhab{\V{I}}{\Set}) : \Set \where \\
\;\;\ldots \\
\\
\idescop{\_\:}{}{} : \PI{\V{I}}{\Set} \IDesc{\V{I}} \To (\V{I} \To \Set) \To \Set    \\
\ldots
}\]

Note that $I \To \IDesc{\V{I}}$ is isomorphic, up to interpretation, to an
endofunctor on $\Set^{\V{I}}$. Indeed, $\V{I} \To \IDesc{\V{I}}$ is interpreted as
$\V{I} \To (\V{I} \To \Set) \To \Set$, which is isomorphic to $(\V{I} \To \Set)
\To (\V{I} \To \Set)$. Fix-point of inductive families is defined over
these indexed functors, hence computing a fix-point of the entire
\emph{family} of functors:

\[\stkl{
\Rule{\stk{ 
          \Gamma \vdash \Bhab{\V{I}}{\Set} \\
          \Gamma \vdash \Bhab{\V{R}}{\V{I} \To \IDesc{\V{I}}} \\
          \Gamma \vdash \Bhab{\V{i}}{\V{I}}}}
     {\Gamma \vdash \Bhab{\IMu{\V{I}}{\V{R}}{\V{i}}}{\Set}} \qquad
\Rule{\stk{
          \Gamma \vdash \Bhab{\V{I}}{\Set} \\
          \Gamma \vdash \Bhab{\V{R}}{\V{I} \To \IDesc{\V{I}}} \\
          \Gamma \vdash \Bhab{\V{i}}{\V{I}} \\
          \Gamma \vdash \Bhab{\V{x}}{\idescop{\V{R}\: \V{i}}{\V{I}}{(\IMu{\V{I}}{\V{R}}{})}}}}
     {\Gamma \vdash \Bhab{\Con{\V{x}}}{\IMu{\V{I}}{\V{R}}{\V{i}}}}
}\]

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \IDesc{\!} (\Bhab{\V{I}}{\Set}) : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DVar{\!}         & \V{I} \To \IDesc{\V{I}}                                       \\
    \DConst{\!}       & \Set \To \IDesc{\V{I}}                                        \\
    \DProd{\!}{\!}    & \PI{\V{\V{D}}, \V{\V{D'}}}{\IDesc{\V{I}}} \IDesc{\V{I}}       \\
    \DSigma{\!}{\!}   & \PI{\V{S}}{\Set} \PIS{\V{S} \To \IDesc{\V{I}}} \IDesc{\V{I}}  \\
    \DPi{\!}{\!}      & \PI{\V{S}}{\Set} \PIS{\V{S} \To \IDesc{\V{I}}} \IDesc{\V{I}} 
\end{array}
}
\vspace{0.2in}
\\
\stk{
\idescop{\_\:}{}{} :_{\PI{\V{I}}{\Set}} \IDesc{\V{I}} \To (\V{\V{I}} \To \Set) \To \Set                  \\
\begin{array}{@{}l@{\:=\:\:}ll}
\idescop{\DVar{\V{i}}}{\V{I}}{\V{P}}      &  \V{P}~\V{i}                                                 \\
\idescop{\DConst{\V{X}}}{\V{I}}{\V{P}}    &  \V{X}                                                   \\
\idescop{\DProd{\V{D}}{\V{D'}}}{\V{I}}{\V{P}} &  \TIMES{\idescop{\V{D}}{\V{I}}{\V{P}}}{\idescop{\V{D'}}{\V{I}}{\V{P}}}       \\
\idescop{\DSigma{\V{S}}{\V{D}}}{\V{I}}{\V{P}} &  \SIGMA{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{P}}                    \\
\idescop{\DPi{\V{S}}{\V{D}}}{\V{I}}{\V{P}}    &  \PI{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{P}}            
\end{array}
}
\end{array}
}\]

@@

\subsection{Tagged indexed descriptions}

\begin{wstructure}
<- Tagged constructor choice
    <- Index available when defining the data-type
        -> Can influence the choice of constructors
        -> Dependently-typed data-types
            <- term (index) influence types
    -> Two parts
        <- Always on the menu
            <- E : EnumU 
            <- ED : E -> IDesc I
        <- Index-dependent
            <- F : I -> EnumU 
            <- FD : (i : I) -> spi (F i) (\_ -> IDesc I)
        -> taggedIDesc I == Sigma E ED x Sigma F FD
        -> toIDesc : (I : Set) -> taggedIDesc I -> (I -> IDesc I)
\end{wstructure}

@@

For convenience, we divide a tagged indexed description in two parts:
first, the constructors that do not depend on the index; then, the
constructors that do. The non-dependent part mirrors the definition
for non-indexed descriptions: we are provided a finite choice of
constructors. The index-depend part simply indexes the choice of
constructors by $I$. Hence, by inspecting the index, it is possible to
enable or disable the constructors.

@@

In the case of vectors, for instance, for the index $\NatZero$, we
would only propose the constructor $\ListNil$. Similarly, for
$\NatSuc{n}$, we only propose the constructor $\ListCons{\!}{\!}$. Hence,
we obtain the following definition:

\[\stk{
\VecD : \Set \To \TagIDesc{\Nat} \\
\begin{array}{@{}ll}
\VecD\: \V{X} = \\ \quad\bigRedBracket{ 
            \begin{array}{l}
                \pair{[]}{\_}{} 
                \\\\
                \bigRedBracket{
                \begin{array}{l} 
                  \LAM{\V{n}} \caseB{\V{n}}{\NatZero \mapsto [\VecNil] \\
                                   (\NatSuc{\V{n}}) \mapsto [\VecCons{\!}{\!}]} \\
                  \LAM{\V{n}} \caseB{\V{n}}{\NatZero \mapsto [\DConst{\Unit}] \\
                                   (\NatSuc{\V{n}}) \mapsto [\DSigma{\V{X}}{\DVar{\V{n}}}]}
                \end{array}}
            \end{array}}
\end{array}
}\]

We will see more instances of this pattern in the following.

\subsubsection{Examples}
\label{sec:idesc-examples}

In order to gain some intuition with $\IDesc{\!}$, let us re-implement
the pattern functor of natural numbers. We refer the reader to
Section~\ref{sec:desc-examples} for its incarnation in the universe of
descriptions. The code is the following:

\[\stk{
\NatD : \IDesc{\Unit} \\
\NatD \mapsto \DSigma{(\EnumT{[ \NatZero, \NatSuc{\!} ]})}
                     {[ \DConst{\Unit} \quad (\DVar{\Void}) ]}
}\]

Because $\Nat$ is just an inductive type, we make no use of the index:
it is the trivial inhabitant of $\Unit$. Therefore, the recursive
argument is materialised by $\DVar{\Void}$, where we were using $\DId$
in the $\Desc$ presentation. This transformation applies to all
inductive types: we have not lost in expressive power during the
transition.

A mandatory exercise consists in describing $\IDesc{\!}$ in itself. With
the experience gained in Section~\ref{sec:desc-levitate}, this
levitation does not pose any issue:

Note that this description also lives in $\IDesc{\Unit}$. Indeed,
for any $\V{I}$, $\IDesc{\V{I}}$ is merely an inductive type. As for the
$\Desc$ levitation, the finite function space $[ \ldots ]$ elaborates into a
special purpose $\F{switchID}$ operator.

@@

%% Index mapper (terminology?)
\newcommand{\Val}[1]{\D{Val}~#1}
\newcommand{\Var}[2]{\D{Var}_{#1}~#2}

%% Hutton expressions
\newcommand{\HExprD}{\C{ExprD}}
\newcommand{\HExprAD}{\C{ExprAlwaysD}}
\newcommand{\HExprID}{\C{ExprIndexedD}}
\newcommand{\HExprVarD}[1]{\C{ExprD}_{\C{Var},#1}}
\newcommand{\HExprFreeD}{\C{ExprD}^{\C{Free}}}
\newcommand{\HExprAFreeD}{\C{ExprAlwaysD}^{\C{Free}}}

@@

The function $\Val{}$, used in the definition of $\Eval{}$, simply
maps an object type $ty$ to the corresponding type in the host
language. Hence, the argument of $\Eval{}$ are ensured to be of the
expected type. We assume $\Nat$ and $\Bool$ represent natural numbers
and booleans in the host language. Further, we also assume the
existence of an addition and comparison operator in $\Nat$,
respectively named $\F{plusHost}$ and $\F{leHost}$. We define $\Val{}$
as follow:

@@

In our universe of descriptions, this data-type is represented by a
tagged indexed description. We use the index to carry the type: the
resulting description is indexed by $\Ty$. We observe that some
constructors are ``polymorph'', namely $\Econd{}{}{}$ and $\Eval{}$. On the
other hand, the $\Eplus$ and $\Ele$ constructors are
index-dependent. $\Eplus$ is defined if and only if the result type --
the index -- is $\Enat$, whereas $\Ele$ is defined if and only if the
index is $\Ebool$. The actual code precisely follows this intuition,
as shown in Figure~\ref{fig:hexpr-full}. For brevity, we use an
informal $\case{\ldots}{\ldots}$ notation, simulating a definition by
pattern-matching. Formally, this corresponds to a call to the
$\switch{\!}{\!}{\!}{\!}$ eliminator.

\begin{figure*}

\[\stkc{
\stk{
\HExprD : \TagIDesc{\Ty} \\
\HExprD \mapsto ( \HExprAD , \HExprID ) \\
} \\
\\
\begin{array}{ll}
\stk{
\HExprAD : \ATagIDesc{\Ty} \\
\HExprAD \mapsto \bigRedBracket{
                 \begin{array}{l}
                   \EnumT{[\Eval{}, \Econd{}{}{}]} \red{,} \\
                   \LAM{\V{ty}}
                   \bigRedBracket{
                   \begin{array}{l}
                   \DConst{(\Val{\V{ty}})} \\
                   \DProd{\DVar{\Ebool}}{\DProd{\DVar{\V{ty}}}{\DVar{\V{ty}}}}
                   \end{array}
                   }
                 \end{array}
                 }
\\
\\
} &
\stk{
\HExprID : \ITagIDesc{\Ty} \\
\HExprID \mapsto \bigRedBracket{
                 \begin{array}{l}
                   \LAM{\V{t}y} \case{\V{t}y}{\Enat \To \EnumT{[\Eplus{\!}{\!}\!]} \\ \Ebool \To \EnumT{[\Ele{\!}{\!}\!]}} \red{,} \\
                   \LAM{\_} \DProd{\DVar{\Enat}}{\DVar{\Enat}} 
                   \end{array}
                   }
}
\end{array}
}\]

@@

The type of $\F{eval}$ is strikingly similar to a
catamorphism. Indeed, implementing a single step of evaluation -- the
algebra -- is sufficient, as $\F{cataI}$ \note{We have not defined
  \F{cataI}!} gives, for free, the full evaluator. The implementation
is as follow:

\[\stk{
\evalOne : \PI{\V{ty}}{\Ty} \idescop{\toIDesc{\HExprD}\: \V{ty}}{\Ty}{\V{ty}}
\To {\Val{\V{ty}}} \\
\begin{array}{@{}l@{\:=\:\:}l}
\evalOne\: \_\: \pair{\Eval{}}{\V{x}}{}                                         & \V{x} \\
\evalOne\: \_\: \pair{\Econd{}{}{}}{\pair{\BoolTrue}{\pair{\V{x}}{\_}{}}{}}{}   & \V{x} \\
\evalOne\: \_\: \pair{\Econd{}{}{}}{\pair{\BoolFalse}{\pair{\_}{\V{y}}{}}{}}{}  & \V{x} \\
\evalOne\: \Enat\: \pair{\Eplus{}{}}{\pair{\V{x}}{\V{y}}{}}{}                       & \F{plusHost}\: \V{x}\: \V{y} \\
\evalOne\: \Ebool\: \pair{\Ele{}{}}{\pair{\V{x}}{\V{y}}{}}{}                        & \F{leHost}\: \V{x}\: \V{y} 
\end{array} \\
\\
\evalH : \PI{\V{ty}}{\Ty} 
           \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}} \To
           \Val{\V{ty}} \\
\evalH\: \V{ty}\: \V{term} = \F{cataI}\: \Ty\: 
                                 \toIDesc{\HExprD}\: 
                                 \Val{}\: 
                                 \evalOne\: 
                                 \V{ty}\: 
                                 \V{term}
}\]
\note{pwm: $\evalOne$ rather illustrates my point about the 2 types of
  IFunc floating around, $\idescop{\_}{}{}$ is defined using the 1st
  notion, and used it here as if it were the 2nd. pierre: in this
  case, it was a typo. But you're right, there is two notions and we
  probably want to clarify where these things live.}

@@

However, so far, we are only able to define and manipulate
\emph{closed} terms. By abstracting over $\Val{}$, it is possible to
build and manipulate \emph{open} terms, that is terms with symbolic
variables. On the model of $\Val{}$, we define $\Var{}{}$:

@@

\paragraph{Indexed description:}

>>>>

However, this new class of objects cannot be defined by mere
induction. In the case of vectors, for instance, we have to define the
whole \emph{family} of vector in one go: vectors of all sizes need to
be defined at the same time. Where simply-typed languages have
inductive types, in dependently typed programming the basic grammar of
data-types should be that of inductive families of types. To this end, we
rely on \emph{indexing}. We are going to transform the $\Desc$ universe into
$\IDesc{\!}$, the universe of indexed descriptions. We express inductive families
by the $\V{I} \To \IDesc{\V{I}}$ type.  \note{pwm: Vec might not be the best
motivating example here, since it doesn't appear in the paper (more on that
later). Come to think of it, this is entirely too vague without spelling out
a concrete example.}

\subsection{Desc, atomically}
\label{sec:idesc-atomic-desc}

@@

Before moving on to indexed descriptions, we have to carry out some
maintenance work on descriptions. We presented $\Desc$ as the grammar
of inductive types. Hence, the codes closely follow this grammar. In
the following, we adopt an alternative presentation. With
$\DSigma{\!}{\!}$, we are actually \emph{quoting} a standard
type-former, namely

@@

\[\stk{
\data \IDesc{(\Bhab{\V{I}}{\Set})} : \Set \where \:\ldots \\
\\
\idescop{\_}{}{} : _{\PI{\V{I}}{\Set}} \IDesc{\V{I}} \To (\V{I} \To \Set) \To \Set    \\
\ldots
}\]

Note that $\V{I} \To \IDesc{\V{I}}$ is isomorphic, up to interpretation, to an
endofunctor on $\Set^{\V{I}}$. Indeed, $\V{I} \To \IDesc{\V{I}}$ is interpreted as
$\V{I} \To \Set^{\V{I}} \To \Set$, which is isomorphic to $\Set^{\V{I}}
\To \Set^{\V{I}}$. Fix-point of inductive families is defined over
these indexed functors, hence computing a fix-point of the entire
\emph{family} of functors:

\[\stkl{
\Rule{\Gamma \vdash \Bhab{\V{I}}{\Set} \qquad
      \Gamma \vdash \Bhab{\V{R}}{\V{I} \To \IDesc{\V{I}}}}
     {\Gamma \vdash \Bhab{\IMu{\V{I}}{\V{R}}{\!}}{\V{I}\To\Set}} \qquad
\\:\
\Rule{\begin{array}{l@{\qquad}l}
          \Gamma \vdash \Bhab{\V{I}}{\Set} &
          \Gamma \vdash \Bhab{\V{R}}{\V{I} \To \IDesc{\V{I}}} \\
          \Gamma \vdash \Bhab{\V{i}}{\V{I}} &
          \Gamma \vdash \Bhab{\V{x}}{\idescop{\V{R}\:\V{i}}{\V{I}}{(\IMu{\V{I}}{\V{R}}{\!})}}
      \end{array}}
     {\Gamma \vdash \Bhab{\Con{\V{x}}}{\IMu{\V{I}}{\V{R}}{\V{i}}}}
}\]

@@

\[\stk{
\begin{array}{ll}
\stk{
\data \IDesc{(\Bhab{\V{I}}{\Set})} : \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \DVar{\!}         & \V{I} \To \IDesc{\V{I}}                                       \\
    \DConst{\!}       & \Set \To \IDesc{\V{I}}                                        \\
    \DProd{\!}{\!}    & \PI{\V{\V{D}}, \V{\V{D'}}}{\IDesc{\V{I}}} \IDesc{\V{I}}       \\
    \DSigma{\!}{\!}   & \PI{\V{S}}{\Set} \PIS{\V{S} \To \IDesc{\V{I}}} \IDesc{\V{I}}  \\
    \DPi{\!}{\!}      & \PI{\V{S}}{\Set} \PIS{\V{S} \To \IDesc{\V{I}}} \IDesc{\V{I}} 
\end{array}
}
\vspace{0.2in}
\\
\stk{
\idescop{\_\:}{}{} :_{\PI{\V{I}}{\Set}} \IDesc{\V{I}} \To (\V{\V{I}} \To \Set) \To \Set                  \\
\begin{array}{@{}l@{\:=\:\:}ll}
\idescop{\DVar{\V{i}}}{\V{I}}{\V{P}}      &  \V{P}~\V{i}                                                 \\
\idescop{\DConst{\V{X}}}{\V{I}}{\V{P}}    &  \V{X}                                                   \\
\idescop{\DProd{\V{D}}{\V{D'}}}{\V{I}}{\V{P}} &  \TIMES{\idescop{\V{D}}{\V{I}}{\V{P}}}{\idescop{\V{D'}}{\V{I}}{\V{P}}}       \\
\idescop{\DSigma{\V{S}}{\V{D}}}{\V{I}}{\V{P}} &  \SIGMA{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{P}}                    \\
\idescop{\DPi{\V{S}}{\V{D}}}{\V{I}}{\V{P}}    &  \PI{\V{s}}{\V{S}} \idescop{\V{D}\: \V{s}}{\V{I}}{\V{P}}            
\end{array}
}
\end{array}
}\]

@@

%\subsection{Tagged indexed descriptions}

% This discussion seemed out of place to me, so i've moved it a bit later (pwm).

%\begin{wstructure}
%<- Tagged constructor choice
%    <- Index available when defining the data-type
%        -> Can influence the choice of constructors
%        -> Dependently-typed data-types
%            <- term (index) influence types
%    -> Two parts
%        <- Always on the menu
%            <- E : EnumU 
%            <- ED : E -> IDesc I
%        <- Index-dependent
%            <- F : I -> EnumU 
%            <- FD : (i : I) -> spi (F i) (\_ -> IDesc I)
%        -> taggedIDesc I == Sigma E ED x Sigma F FD
%        -> toIDesc : (I : Set) -> taggedIDesc I -> (I -> IDesc I)
%\end{wstructure}
%
%In Section~\ref{sec:desc-examples}, we have defined a \emph{tagged}
%form for descriptions. This format follows the usual presentation of
%inductive types as sum-of-product. With indexed description, this
%definition can be generalized. Indeed, when defining an indexed
%data-type, we have access to this index. Therefore, we can use this
%index to influence the choice of constructors. This captures the
%essence of dependent data-types: a term -- the index -- has the
%ability to influence the data-type.
%
%For convenience, we divide a tagged indexed description in two parts:
%first, the constructors that do not depend on the index; then, the
%constructors that do. The non-dependent part mirrors the definition
%for non-indexed descriptions: we are provided a finite choice of
%constructors. The index-depend part simply indexes the choice of
%constructors by $I$. Hence, by inspecting the index, it is possible to
%enable or disable the constructors. 
%
%\[\stk{
% \TagIDesc{\V{I}}  \mapsto \TIMES{\ATagIDesc{\V{I}}}{\ITagIDesc{\V{I}}} \\
% \ATagIDesc{\V{I}} \mapsto \SIGMA{\V{E}}{\EnumU} (\PI{\V{i}}{\V{I}} \spi{\V{E}}{(\LAM{\_} \IDesc{\V{I}})}) \\
% \ITagIDesc{\V{I}} \mapsto \\
%\qquad\SIGMA{\V{F}}{\V{I} \To \EnumU} (\PI{\V{i}}{\V{I}} \spi{(\V{F}\: \V{i})}{(\LAM{\_} \IDesc{\V{I}})}) 
%}\]
%
%
%\begin{wstructure}
%<- Vectors
%    Do we treat them in the end? 
%    What can we say here we haven't with Fin?
%\end{wstructure}
%
%In the case of vectors, for instance, for the index $\NatZero$, we
%would only propose the constructor $\ListNil$. Similarly, for
%$\NatSuc{n}$, we only propose the constructor $\ListCons{\!}{\!}$. Hence,
%we obtain the following definition:
%
%\newcommand{\VecD}{\C{VecD}}
%\newcommand{\VecNil}{\etag{\CN{vnil}}}
%\newcommand{\VecCons}[2]{\etag{\CN{vcons}}\:#1\:#2}
%
%\[\stk{
%\VecD : \Set \To \TagIDesc{\Nat} \\
%\begin{array}{@{}ll}
%\VecD\: \V{X} = \\ \quad\bigRedBracket{ 
%            \begin{array}{l}
%                \pair{[]}{\_}{} 
%                \\\\
%                \bigRedBracket{
%                \begin{array}{l} 
%                  \LAM{\V{n}} \caseB{\V{n}}{\NatZero \mapsto [\VecNil] \\
%                                   (\NatSuc{\V{n}}) \mapsto [\VecCons{\!}{\!}]} \\
%                  \LAM{\V{n}} \caseB{\V{n}}{\NatZero \mapsto [\DConst{\Unit}] \\
%                                   (\NatSuc{\V{n}}) \mapsto [\DSigma{\V{X}}{\DVar{\V{n}}}]}
%                \end{array}}
%            \end{array}}
%\end{array}
%}\]
%
%We will see more instances of this pattern in the following.

\subsection{Examples}
\label{sec:idesc-examples}

In order to gain some intuition with $\IDesc{\!}$, let us re-implement
the pattern functor of natural numbers. This follows the implementation of
$\NatD$ from \ref{sec:idesc-atomic-desc}:

\[\stk{
\NatD : \IDesc{\Unit} \\
\NatD \mapsto \DSigma{(\EnumT{[ \NatZero, \NatSuc{\!} ]})}
                     {[ (\DConst{\Unit}) \quad (\DVar{\Void}) ]}
}\]

Because $\Nat$ is just an inductive type, $\NatD$ represents a
$\Unit$-indexed functor. Therefore, the recursive
argument is materialised by $\DVar{\Void}$, where we were using $\DId$
in the $\Desc$ presentation. Indeed, all inductive types can be
translated into $\IDesc{\!}$ in exactly this way.

If we note that, for any $\V{I}$, $\IDesc{\V{I}}$ itself is merely an inductive 
\emph{type} then we can certainly define its encoding in $\IDesc{\Unit}$:

Using this description we can repeat the act of levitation 
from section~\ref{sec:desc-levitate}. As before we rely on a special purpose 
$\F{switchID}$ operator to build the finite function space
$\bigRedBracket{\ldots}$.

\paragraph{Vectors:}

So far, the examples we have seen lives in $\IDesc{\Unit}$, hence are
not using any indexing. We remedy this by encoding the vectors:

Recall that the constructors of $\D{Vec}$. The constructors $\VecNil$ and
$\VecCons$ are only defined for an index $\NatZero$ and $\NatSuc$ 
respectively.:

\[
\stk{
\data \D{Vec}\: \PITEL{\V{X}}{\Set} : \PI{\V{n}}{\Nat} \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \VecNil      & \D{Vec}\:\V{X}\:{\NatZero)}   \\
    \VecCons     & _{\PI{\V{n}}{\Nat}}\V{X} \To \D{Vec}\:{\V{X}}\:{\V{n}} \To \D{Vec}\:{\V{X}}\:{(\NatSuc{\V{n}})}
\end{array}
}
\]

\[\stk{
\VecD : \Set \To \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}ll}
\VecD\: \V{X}\: \V{n} = \DSigma{\!}{\!} \!\!\!\!& (\EnumT{[ \VecNil , \VecCons{\!}{\!} ]}) \\
                        & \bigRedBracket{
                          \begin{array}{l}
                            \DConst{(\V{n} \PropEq \NatZero)} \\
                            \DSigma{\Nat}{\LAM{\V{m}} \stk{\DProd{\DVar{\V{m}}}{\\\DConst{(\V{n} \PropEq \NatSuc{\V{m}})}}}}
                          \end{array}
                          }
\end{array}
}\]

If the $\VecNil$ tag is chosen, a proof must be provided that the index is equal
to $\NatZero$. In the $\VecCons$ case, we first store an
element $\V{m}$ of $\Nat$, with a $\DSigma{\!}{\!}$. However, the
constraint stipulates that $\V{m}$ cannot be \emph{any} natural
numbers: it must be ``the index minus one''. This translates into the
constraint $\V{n} \PropEq \NatSuc{\V{m}}$.

However, one could, quite rightfully, be worried by the introduction
of equalities in the picture. In a programming context, equalities
haunt us, begging for proofs. Hence, whereas the developer typed a
constructor-constrained data-type, the elimination form will involve
equalities and their proofs. As much as possible, one would rather
avoid suffering from this burden. However, the equalities introduced
here are rather special: they are \emph{equations} constraining
variables to certain terms. \citet{goguen:pattern-matching} have
shown, in the context of dependent pattern matching, how to tame these
equations.

Further, we can actually remove these equations altogether. Let us
look back at $\D{Vec}$. We note that the equations are introduced
because we are \emph{storing} the index of the inductive family.
However, \emph{inductive families need not store their
  indices}~\cite{brady:index-inductive-families}.  By examining the
incoming index, we can apply the \emph{forcing} and \emph{de-tagging}
optimisations to our initial definition of $\D{Vec}$ gives the
following, equivalent definition:

\[\stk{
\VecD \:\PITEL{\V{X}}{\Set} : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}lll}
\VecD\:\V{X}\: \NatZero     & = & \DConst{\Unit} \\
\VecD\:\V{X}\: (\NatSuc{\V{n}}) & = & \DProd{\DConst{\V{X}}}{\DVar{\V{n}}}
\end{array}

}\]

The equations (and constructors) have simply disappeared. Note that it is
straightforward to turn this definition into a tagged form, where in each
case, only one constructor choice is available to the developer. In another
example, $\Fin{\!}$, we can force, but not detag, since both $\FinZero$ and 
$\FinSuc{\!}$ both target $\NatSuc{\!}$:

\[
\stk{
\data \Fin{\!} : \PI{\V{n}}{\Nat} \Set \where \\
\;\;\begin{array}{@{}l@{\::\:\:}l@{\quad}l}
    \FinZero      & _{\PI{\V{n}}{\Nat}}\Fin{(\NatSuc{\V{n}})}   \\
    \FinSuc{\!}   & _{\PI{\V{n}}{\Nat}}\Fin{\V{n}} \To \Fin{(\NatSuc{\V{n}})}
\end{array}
\vspace{0.1in}
\\
\FinD : \Nat \To \IDesc{\Nat} \\
\begin{array}{@{}llll}
\FinD\: \NatZero     & = & \multicolumn{2}{l}{\DSigma{(\EnumT{[]})}{\Void} } \\
\FinD\: (\NatSuc{\V{n}}) & = & \DSigma{\!}{\!}\!\!\!\!\! & (\EnumT{[ \FinZero , \FinSuc{\!} ]}) \\
                     &   &             & \bigRedBracket{
                                         \begin{array}{l}
                                         \DConst{\Unit} \\
                                         \DVar{\V{n}}
                                         \end{array}
                                         }
\end{array}
}\]

We should
precise that forcing a description is not guaranteed to remove all
constraints. It is subject to future work to see if constraints can be
entirely eradicated, or presented more conveniently to the developer.
Finally, it is worth mentioning that these optimisations are 
\emph{source-to-source} transformations on descriptions.

\subsubsection{Tagged Indexed Descriptions}

@@

For convenience, we divide a tagged indexed description in two parts:
first, the constructors that do not depend on the index; then, the
constructors that do. The non-dependent part mirrors the definition
for non-indexed descriptions: we are provided a finite choice of
constructors. The index-depend part simply indexes the choice of
constructors by $\V{I}$. Hence, by inspecting the index, it is
possible to enable or disable the constructors.

@@

In the case of a tagged $\D{Vec}$, for instance, for the index $\NatZero$, we
would only propose the constructor $\ListNil$. Similarly, for
$\NatSuc{n}$, we would only propose the constructor $\ListCons{\!}{\!}$. We
will see a concrete example of this in the following example.

@@

%% Index mapper (terminology?)
\newcommand{\Val}[1]{\F{Val}~#1}
\newcommand{\Var}[2]{\F{Var}_{#1}~#2}

%% Hutton expressions
\newcommand{\HExprD}{\C{ExprD}}
\newcommand{\HExprAD}{\C{ExprAD}}
\newcommand{\HExprID}{\C{ExprID}}
\newcommand{\HExprVarD}[1]{\C{ExprD}_{\F{Var},#1}}
\newcommand{\HExprFreeD}{\C{ExprD}^{\C{Free}}}
\newcommand{\HExprAFreeD}{\C{ExprAD}^{\C{Free}}}

@@

The function $\Val{}$, used in the definition of $\Eval{}$, simply
maps an object type $ty$ to the corresponding type in the host
language. Hence, the argument of $\Eval{}$ are ensured to be of the
expected type. We assume $\Nat$ and $\Bool$ represent natural numbers
and booleans in the host language. Further, we also assume the
existence of an addition and comparison operator in $\Nat$,
respectively named $\F{plusHost}$ and $\F{leHost}$. We define $\Val{}$
as follows:

@@

In our universe of descriptions, this data-type is represented by a
tagged indexed description. We use the index to carry the type: the
resulting description is indexed by $\Ty$. We observe that some
constructors are ``polymorphic'', namely $\Econd{}{}{}$ and $\Eval{}$. On the
other hand, the $\Eplus$ and $\Ele$ constructors are
index-dependent. $\Eplus$ is defined if and only if the result type --
the index -- is $\Enat$, whereas $\Ele$ is defined if and only if the
index is $\Ebool$. The actual code precisely follows this intuition,
as shown in Figure~\ref{fig:hexpr-full}. For brevity, we use an
informal $\case{\ldots}{\ldots}$ notation, simulating a definition by
pattern-matching. Formally, this corresponds to a call to the
$\switch{\!}{\!}{\!}{\!}$ eliminator.

\[\stk{
\stk{
\HExprD : \TagIDesc{\Ty} \\
\HExprD \mapsto ( \HExprAD , \HExprID ) \\
} \\
\\
\stk{
\HExprAD : \ATagIDesc{\Ty} \\
\HExprAD \mapsto \bigRedBracket{
                 \begin{array}{l}
                   {[\Eval{}, \Econd{\!}{\!}{\!}\!\,]} \red{,} \\
                   \LAM{\V{ty}}
                   \bigRedBracket{
                   \begin{array}{l}
                   \DConst{(\Val{\V{ty}})} \\
                   \DProd{\DVar{\Ebool}}{\DProd{\DVar{\V{ty}}}{\DVar{\V{ty}}}}
                   \end{array}
                   }
                 \end{array}
                 }
\\
\\
} 
\\
\stk{
\HExprID : \ITagIDesc{\Ty} \\
\HExprID \mapsto \bigRedBracket{
                 \begin{array}{l}
                   \LAM{\V{ty}} \case{\V{ty}}{\Enat \To {[\Eplus{\!}{\!}\!\,]} \\ \Ebool \To {[\Ele{\!}{\!}\!\,]}} \red{,} \\
                   \LAM{\_} \DProd{\DVar{\Enat}}{\DVar{\Enat}} 
                   \end{array}
                   }
}
}\]

@@

Where we overload the notation $\toIDesc{\V{TID}}$ to denote the
indexed description computed from the tagged indexed description
$\V{TID}$.

The type of $\F{eval}$ is strikingly similar to a
catamorphism. Indeed, implementing a single step of evaluation -- the
algebra -- is sufficient, as $\F{cataI}$ \note{We have not defined
  \F{cataI}!} gives, for free, the full evaluator. The implementation
is as follows:

\[\stk{
\evalOne : \PI{\V{ty}}{\Ty} \idescop{\toIDesc{\HExprD}\:\V{ty}}{\Ty}{\Val{\!}}
\To {\Val{\V{ty}}} \\
\begin{array}{@{}l@{\:=\:\:}l}
\evalOne\: \_\: \pair{\Eval{}}{\V{x}}{}                                         & \V{x} \\
\evalOne\: \_\: \pair{\Econd{}{}{}}{\pair{\BoolTrue}{\pair{\V{x}}{\_}{}}{}}{}   & \V{x} \\
\evalOne\: \_\: \pair{\Econd{}{}{}}{\pair{\BoolFalse}{\pair{\_}{\V{y}}{}}{}}{}  & \V{x} \\
\evalOne\: \Enat\: \pair{\Eplus{}{}}{\pair{\V{x}}{\V{y}}{}}{}                       & \F{plusHost}\: \V{x}\: \V{y} \\
\evalOne\: \Ebool\: \pair{\Ele{}{}}{\pair{\V{x}}{\V{y}}{}}{}                        & \F{leHost}\: \V{x}\: \V{y} 
\end{array} \\
\\
\evalH : \PI{\V{ty}}{\Ty} 
           \IMu{\Ty}{\toIDesc{\HExprD}}{\V{ty}} \To
           \Val{\V{ty}} \\
\evalH\: \V{ty}\: \V{term} = \F{cataI}\: \Ty\: 
                                 \toIDesc{\HExprD}\: 
                                 \Val{}\: 
                                 \evalOne\: 
                                 \V{ty}\: 
                                 \V{term}
}\]
\note{pwm: $\evalOne$ rather illustrates my point about the 2 types of
  IFunc floating around, $\idescop{\_}{}{}$ is defined using the 1st
  notion, and used it here as if it were the 2nd. pierre: in this
  case, it was a typo. But you're right, there is two notions and we
  probably want to clarify where these things live.}

@@

However, so far, we are only able to define and manipulate
\emph{closed} terms. By abstracting over $\Val{}$, it is possible to
build and manipulate \emph{open} terms, that is, terms with symbolic
variables. On the model of $\Val{}$, we define $\Var{}{}$:

@@
